---
title: "class12"
author: "Nicolas Song"
format: pdf
---

# Import countData and colData

```{r}
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <- read.csv("airway_metadata.csv")
```

```{r}
head(counts)
```

```{r}
head(metadata)
```

**Q1. How many genes are in this dataset?**

There are 38694 genes in this dataset.

```{r}
# Seeing how many genes
nrow(counts)
```

**Q2. How many ‘control’ cell lines do we have?**

There are 4 control cell lines.

```{r}
# Seeing how many of each cell line
table(metadata$dex)
```

# Toy differential gene expression

```{r}
control <- metadata[metadata[,"dex"]=="control",]

# Finding sample ID for treated
control.counts <- counts[,control$id]

# Calculating mean counts per gene for control samples
control.mean <- rowSums(control.counts )/4 

# Inspecting results
head(control.mean)
```

**Which of these methods do you prefer and why?**

I like the base R method better because it feels more straight-forward.

**Q3. How would you make the above code in either approach more robust? Is there a function that could help here?**

Using nrow(control) instead of 4 would make it so that if more samples were added, the correct number of genes would be input into the code since the number of genes would equal the number of controls.

**Q4. Follow the same procedure for the treated samples (i.e. calculate the mean per gene across drug treated samples and assign to a labeled vector called treated.mean)**

```{r}
treated <- metadata[metadata[,"dex"]=="treated",]

# Finding sample ID for treated
treated.counts <- counts[,treated$id]

# Calculating mean counts per gene for control samples
treated.mean <- rowSums(treated.counts )/4 

# Inspecting results
head(treated.mean)
```

```{r}
# Combining meancount data
meancounts <- data.frame(control.mean, treated.mean)

# Sum of mean counts across all genes for each group
colSums(meancounts)
```

**Q5 (a). Create a scatter plot showing the mean of the treated samples against the mean of the control samples.**

```{r}
plot(meancounts[,1],meancounts[,2], xlab="Control", ylab="Treated")
```

**Q5 (b).You could also use the ggplot2 package to make this figure producing the plot below. What geom_?() function would you use for this plot?**

geom_point()

```{r}
library(ggplot2)
ggplot(meancounts, aes(control.mean, treated.mean)) + geom_point()
```

**Q6. Try plotting both axes on a log scale. What is the argument to plot() that allows you to do this?**

log

```{r}
plot(meancounts[,1],meancounts[,2], xlab="Control", ylab="Treated", 
 # Putting both axes on a log scale    
     log="xy")
```

```{r}
# Calculating log2 of the fold change
meancounts$log2fc <- 
  log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])

# Inspecting results
head(meancounts)
```

```{r}
# Filtering data
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)
to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

**Q7. What is the purpose of the arr.ind argument in the which() function call above? Why would we then take the first column of the output and need to call the unique() function?**

arr.ind makes it so that which() returns values in the rows and columns that have a zero count. Calling unique() on zero.vals makes it so that rows that have such a zero count in multiple samples are not counted multiple times.

```{r}
# Filtering for up-regulated genes
up.ind <- mycounts$log2fc > 2

# Filtering for down-regulated genes
down.ind <- mycounts$log2fc < (-2)
```

**Q8. Using the up.ind vector above can you determine how many up regulated genes we have at the greater than 2 fc level?**

250 up-regulated genes greater than 2 fc. 

```{r}
sum(up.ind==TRUE)
```

**Q9. Using the down.ind vector above can you determine how many down regulated genes we have at the greater than 2 fc level?**

367 down-regulated genes greater than 2 fc.

```{r}
sum(down.ind==TRUE)
```

**Q10. Do you trust these results? Why or why not?**

I do not fully trust these results because there has not been any procedure to test for statistical significance of these differences.

# Importing data

```{r}
library(DESeq2)
citation("DESeq2")
```

```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex)
dds
```

# Principal Component Analysis (PCA)

```{r}
# Variance stabilizing transformation
vsd <- vst(dds, blind = FALSE)

# Plotting PCs
plotPCA(vsd, intgroup = c("dex"))
```

```{r}
pcaData <- plotPCA(vsd, intgroup=c("dex"), returnData=TRUE)
head(pcaData)
```

```{r}
# Calculate percent variance per PC for the plot axis labels
percentVar <- round(100 * attr(pcaData, "percentVar"))
```

```{r}
# Plotting using ggplot
ggplot(pcaData) +
  aes(x = PC1, y = PC2, color = dex) +
  geom_point(size =3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  theme_bw()
```

# DESeq analysis

```{r}
dds <- DESeq(dds)
```

# Getting results

```{r}
res <- results(dds)
head(res)
```

```{r}
head(as.data.frame(res))
```

```{r}
summary(res)
```

```{r}
res05 <- results(dds, alpha=0.05)
summary(res05)
```

# Adding annotation data

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
columns(org.Hs.eg.db)
```

```{r}
res$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     # Our genenames
                     keytype="ENSEMBL",
                     # The format of our genenames
                     column="SYMBOL",
                     # The new format we want to add
                     multiVals="first")

head(res)
```

**Q11. Run the mapIds() function two more times to add the Entrez ID and UniProt accession and GENENAME as new columns called res$entrez, res$uniprot and res$genename.**

```{r}
# Creating new columns using mapIds()
res$entrez <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     column="ENTREZID",
                     keytype="ENSEMBL",
                     multiVals="first")

res$uniprot <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     column="UNIPROT",
                     keytype="ENSEMBL",
                     multiVals="first")

res$genename <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     column="GENENAME",
                     keytype="ENSEMBL",
                     multiVals="first")

head(res)
```

```{r}
ord <- order( res$padj )
#View(res[ord,])
head(res[ord,])
```

```{r}
write.csv(res[ord,], "deseq_results.csv")
```

# Volcano Plots

```{r}
plot( res$log2FoldChange,  -log(res$padj), 
      xlab="Log2(FoldChange)",
      ylab="-Log(P-value)")
```

```{r}
plot( res$log2FoldChange,  -log(res$padj), 
 ylab="-Log(P-value)", xlab="Log2(FoldChange)")

# Add some cut-off lines
abline(v=c(-2,2), col="darkgray", lty=2)
abline(h=-log(0.05), col="darkgray", lty=2)
```

```{r}
# Setup our custom point color vector 
mycols <- rep("gray", nrow(res))
mycols[ abs(res$log2FoldChange) > 2 ]  <- "red" 

inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

# Volcano plot with custom colors 
plot( res$log2FoldChange,  -log(res$padj), 
 col=mycols, ylab="-Log(P-value)", xlab="Log2(FoldChange)" )

# Cut-off lines
abline(v=c(-2,2), col="gray", lty=2)
abline(h=-log(0.1), col="gray", lty=2)
```

```{r}
library(EnhancedVolcano)
x <- as.data.frame(res)

EnhancedVolcano(x,
    lab = x$symbol,
    x = 'log2FoldChange',
    y = 'pvalue')
```

